---
title: "Profile"
author: "Jack Scanlon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# The `SAILR::Profile` class

This `R6` class is intended to be used as a secret manager for SAIL DB authentication; it's primarily intended to be used internally by [SAILR::Connection](./Connection.html) but it has been exported for use as it might be useful for your other projects.

If you're only interested in using the SAIL database interface, _i.e._ the `SAILR::Connection` class, then you can safely ignore this documentation!

## Getting Started

> ***Note:** This is a lightweight `R6` wrapper around the `Keyring` package +/- changes to improve accessibility & ease of use within SAIL Databank. The `Keyring` package is available on [Cran](https://cran.r-project.org/web/packages/keyring/index.html) and both authored and maintained by Gabor Csardi. Github repository is available outside SAIL's Gateway TRE [here](https://github.com/r-lib/keyring/)*

### Creating a `SAILR::Profile` instance

> ***Note:** Don't forget to load the package into your enviroment by entering the following: `library(SAILR)`*  

You can create a `Profile` instance by entering the following:

```{r eval=FALSE}
profile = Profile$new(
  # Sets the keychain name, i.e. the name of the keychain that contains
  # the secrets defined by the profile
  #
  #  NOTE:
  #   - Defaults to `SAILDB` but can be changed to any other name if you'd prefer
  #     to separate your secrets between applications
  #
  keychain.name = 'SAILDB'
)
```

Or, you could grab your `Profile` instance from an active `SAILR::Connection` like so:

```{r eval=FALSE}
db = Connection$new() # if you haven't already connected

# Get the active profile (if any)
profile = db$get.profile()
```

## Example usage

Please see example usage below:

```{r eval=FALSE}
profile = Profile$new() # if you haven't already got a profile instance

# Attempt to check whether we have a secret for a username
# associated with a database
#
# NOTE: the database argument defaults to `PR_SAIL`
#
profile$has.secrets('some_username', 'some_database')

# Attempts to collect the secrets associated with the specified
# username and database
#
# NOTE: the database argument defaults to `PR_SAIL`
#
profile$get.secrets('some_username', 'some_database')

# Checks whether the 'super.secret.password' is the secret stored within
# the profile for the 'some_username' account for the 'some_database' database
#
# NOTE: the database argument defaults to `PR_SAIL`
#
profile$is.secret('some_username', 'super.secret.password', 'some_database')

# Sets or creates & stores the secret for the 'some_username' account
# with the 'super.secret.password' secret, associated with the 'some_database' database
#
# NOTE: the database argument defaults to `PR_SAIL`
#
profile$is.secret('some_username', 'super.secret.password', 'some_database')

# Attempts to remove the 'some_username' secret associated with the 'some_database' database
#
# NOTE: the database argument defaults to `PR_SAIL`
#
profile$get.secrets('some_username', 'some_database')

# Field members...

## Get the current keychain name associated with this profile
profile$keychain.name

## Get the username of the account associated with this machine
profile$system.user

# Static methods...

## Check if some object is a `SAILR::Profile`
##
##  NOTE:
##   - Static methods CANNOT be accessed via the `profile` instance;
##     they can only be accessed via the top-level class like below:
##
if (Profile$is(profile)) {
  print('It is a Profile!')
}

try({
  profile$is(profile) # this will fail!
})

## Attempt to clear a secret associated with a username and database
##
##  NOTE:
##   - As above, this can only be accessed by the top-level class
##   - The latter argument defaults to `SAILDB` and describes the keychain name
##
Profile$clear('PR_SAIL:some_username', 'SAILDB')
```
